---
title: "Building Fantasy-Forward Park Factors from Statcast Batted-Ball Residuals"
author: "C. Kaufman + Codex"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: false
    theme: cosmo
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6, dpi = 140)
suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
  library(ggplot2)
  library(knitr)
})

root <- normalizePath(file.path(getwd(), "data", "processed", "park_factors"), mustWork = TRUE)

overall <- read_csv(file.path(root, "park_factors_savant_style_clean_2026.csv"), show_col_types = FALSE)
h1 <- read_csv(file.path(root, "park_factors_savant_style_clean_2026_1H.csv"), show_col_types = FALSE)
h2 <- read_csv(file.path(root, "park_factors_savant_style_clean_2026_2H.csv"), show_col_types = FALSE)
known <- read_csv(file.path(root, "park_factor_known_effects_2026.csv"), show_col_types = FALSE)
val_sum <- read_csv(file.path(root, "validation_summary.csv"), show_col_types = FALSE)
val_det <- read_csv(file.path(root, "validation_detail.csv"), show_col_types = FALSE)
weights <- read_csv(file.path(root, "park_factors_savant_style_weights.csv"), show_col_types = FALSE)
invariance <- read_csv(file.path(root, "invariance_checks.csv"), show_col_types = FALSE)
meta <- read_csv(file.path(root, "run_metadata.csv"), show_col_types = FALSE)

weighted_mean <- function(x, w) {
  ok <- is.finite(x) & is.finite(w)
  if (!any(ok)) return(NA_real_)
  sum(x[ok] * w[ok]) / sum(w[ok])
}

rows_modeled <- as.numeric(meta$value[meta$key == "rows_modeled"]) %>% as.integer()
seasons_modeled <- meta$value[meta$key == "seasons_modeled"]
exclude_2020 <- meta$value[meta$key == "exclude_seasons"]
mean_rmse_model <- weighted_mean(val_sum$rmse_model, val_sum$n_park_half)
mean_rmse_zero <- weighted_mean(val_sum$rmse_zero, val_sum$n_park_half)
mean_rmse_prev <- weighted_mean(val_sum$rmse_prev, val_sum$n_park_half)
mean_corr <- weighted_mean(val_sum$corr_model_vs_realized, val_sum$n_park_half)
mean_corr_prev <- weighted_mean(val_sum$corr_prev_vs_realized, val_sum$n_park_half)
mean_slope <- weighted_mean(val_sum$calibration_slope, val_sum$n_park_half)

def_corr <- invariance$value[invariance$metric == "corr_park_effect_vs_home_team_defense"]
off_corr <- invariance$value[invariance$metric == "corr_park_effect_vs_home_team_xwoba_con"]

w_tbl <- weights %>% mutate(weight = round(weight, 2))

overall_tbl <- overall %>%
  mutate(`Total BBE` = as.integer(round(`Total BBE`, 0))) %>%
  arrange(desc(`Overall Park Factor`))

h_split <- h1 %>%
  select(Team, Park, `Overall Park Factor`) %>%
  rename(`1H Overall PF` = `Overall Park Factor`) %>%
  inner_join(
    h2 %>% select(Team, Park, `Overall Park Factor`) %>% rename(`2H Overall PF` = `Overall Park Factor`),
    by = c("Team", "Park")
  ) %>%
  mutate(`2H-1H` = `2H Overall PF` - `1H Overall PF`) %>%
  arrange(desc(abs(`2H-1H`)))

fit_vs_realized <- val_det %>%
  mutate(season = as.factor(season))
```

## Why Build a New Park Factor?

For fantasy pitcher streaming, the main park-risk pathways are not identical to public one-number park factors. The specific risks we care about are:

- Hit environment (BACON / contact-hit environment)
- HR environment
- Overall run-scoring environment

This build uses event-level Statcast batted-ball data and explicitly removes team quality effects (batter, pitcher, batting team, fielding team, and fielding quality) so we isolate park signal.

## Data Scope

- Statcast BBE seasons modeled: `r seasons_modeled`
- Excluded season: `r exclude_2020`
- Total BBE rows modeled: `r format(rows_modeled, big.mark = ',')`
- Park-era handling: explicit team/venue era map (`park_era_events.csv`) to account for moves and wall changes
- 2026 output filter: only parks expected to be active in 2026

## Model Design

### Core Residual Target

Primary target at the event level:

\[
\text{resid} = \text{wOBAcon} - \text{xwOBAcon}
\]

Core component residuals:

- `bacon_resid = hit_on_contact - xBA_on_contact`
- `hr_resid = HR_on_contact - xHR_on_contact`
- `xbh_resid = XBH_on_contact - xXBH_on_contact`

Where xBA is taken from `estimated_ba_using_speedangle` when available, with contact-shape smoothing fallback by EV/LA/stand bins.

### Hierarchical Structure

Random intercepts:

- `park_era_id`
- `park_era_half_id`
- `batter_season_id`
- `pitcher_season_id`
- `fielding_team_season_id`
- `batting_team_season_id`

Fixed effects include:

- `half` (1H: Mar-Jun, 2H: Jul-Sep)
- weather/drag features when available
- defense composite (z-scored OAA, DRS, UZR mix; falls back to DRS+UZR when OAA missing)

Measurement-era term (Trackman vs Hawk-Eye) was tested and left OFF in the final run because it did not improve holdout performance in this dataset.

## Chosen Fantasy Weights

Final displayed index uses your chosen blend:

```{r weights-table}
kable(w_tbl, align = "lc", caption = "User-selected component weights for Overall Park Factor index")
```

## Validation

Validation is rolling out-of-sample by season (train previous 3 years, test next season). Holdout unit is park-half.

### Headline Validation Numbers

- Weighted mean RMSE (model): `r sprintf('%.4f', mean_rmse_model)`
- Weighted mean RMSE (zero baseline): `r sprintf('%.4f', mean_rmse_zero)`
- Weighted mean RMSE (previous-years park baseline): `r sprintf('%.4f', mean_rmse_prev)`
- Weighted mean correlation (predicted vs realized): `r sprintf('%.3f', mean_corr)`
- Baseline correlation (previous-years park mean): `r sprintf('%.3f', mean_corr_prev)`
- Weighted mean calibration slope: `r sprintf('%.3f', mean_slope)`

```{r validation-rmse-plot}
val_long <- bind_rows(
  val_sum %>% transmute(season, model = "Current model", rmse = rmse_model),
  val_sum %>% transmute(season, model = "Zero baseline", rmse = rmse_zero),
  val_sum %>% transmute(season, model = "Previous park mean", rmse = rmse_prev)
)

ggplot(val_long, aes(x = season, y = rmse, color = model)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Current model" = "#1b9e77", "Zero baseline" = "#d95f02", "Previous park mean" = "#7570b3")) +
  labs(
    title = "Rolling Holdout RMSE by Season",
    x = "Holdout season",
    y = "RMSE",
    color = ""
  ) +
  theme_minimal(base_size = 12)
```

```{r validation-scatter}
set.seed(42)
plot_dat <- fit_vs_realized
if (nrow(plot_dat) > 3000) {
  plot_dat <- plot_dat %>% sample_n(3000)
}

ggplot(plot_dat, aes(x = pred_effect, y = realized)) +
  geom_point(alpha = 0.35, size = 1.2, color = "#2c7fb8") +
  geom_smooth(method = "lm", se = FALSE, color = "#d95f02", linewidth = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "#666666") +
  labs(
    title = "Holdout Park-Half Fit: Predicted vs Realized Residual",
    x = "Predicted park-half residual",
    y = "Realized park-half residual"
  ) +
  theme_minimal(base_size = 12)
```

## Stability and Invariance Checks

Goal: park factors should be agnostic to roster quality.

- Corr(park effect, home-team xwOBAcon): `r sprintf('%.3f', off_corr)`
- Corr(park effect, home-team defense composite): `r sprintf('%.3f', def_corr)`

Both are near zero, consistent with successful decontamination of team quality.

## 2026 Overall Park Factor Snapshot

```{r overall-top-bottom-table}
show_top <- overall_tbl %>% slice_head(n = 10)
show_bottom <- overall_tbl %>% arrange(`Overall Park Factor`) %>% slice_head(n = 10)

kable(show_top, caption = "Top 10 hitter-friendly parks (2026 usable parks)")
kable(show_bottom, caption = "Top 10 pitcher-friendly parks (2026 usable parks)")
```

```{r overall-bar-plot}
plot_overall <- overall_tbl %>%
  mutate(ParkLabel = paste0(Team, " - ", Park)) %>%
  arrange(`Overall Park Factor`) %>%
  mutate(ParkLabel = factor(ParkLabel, levels = ParkLabel))

ggplot(plot_overall, aes(x = ParkLabel, y = `Overall Park Factor`, fill = `Overall Park Factor` >= 100)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#d7301f", "FALSE" = "#2b8cbe"), guide = "none") +
  geom_hline(yintercept = 100, linetype = 2, color = "#444444") +
  labs(
    title = "Overall Park Factor Index (100 = league average)",
    x = "",
    y = "Overall PF Index"
  ) +
  theme_minimal(base_size = 11)
```

## Known Park Effects (Useful for Streaming)

```{r hr-bacon-gap-plot}
hb <- known %>% filter(Analysis == "HR vs BACON Gap")

if (nrow(hb) > 0) {
  ggplot(hb, aes(x = reorder(paste(Team, Park, sep = " - "), Difference), y = Difference, fill = Difference > 0)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("TRUE" = "#d7301f", "FALSE" = "#2b8cbe"), guide = "none") +
    geom_hline(yintercept = 0, linetype = 2, color = "#444444") +
    labs(
      title = "HR PF minus BACON PF (positive = more HR-friendly than hit-friendly)",
      x = "",
      y = "PF difference"
    ) +
    theme_minimal(base_size = 11)
}
```

```{r half-split-plot}
plot_half <- h_split %>% slice_head(n = 15) %>%
  mutate(label = paste(Team, Park, sep = " - "))

ggplot(plot_half, aes(x = reorder(label, `2H-1H`), y = `2H-1H`, fill = `2H-1H` > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#d7301f", "FALSE" = "#2b8cbe"), guide = "none") +
  geom_hline(yintercept = 0, linetype = 2, color = "#444444") +
  labs(
    title = "Largest 1H vs 2H Overall PF Movers",
    subtitle = "Positive means park plays more hitter-friendly in 2H",
    x = "",
    y = "2H minus 1H PF"
  ) +
  theme_minimal(base_size = 11)
```

## Comparison to Public Park Factor Frameworks

This model is intentionally different from common public park-factor outputs:

- **Statcast Park Factors (Savant)**: excellent event-quality framework, but published outputs are not directly defense-decontaminated with explicit team-defense controls in this format.
- **FanGraphs Park Factors**: stable and widely used, but generally oriented to run/stat outcomes at aggregate level rather than BBE-level residual decomposition with explicit defense filtering.
- **This model**: event-level residual architecture with mixed-effects controls for batter/pitcher/team quality and explicit defensive composite adjustment.

Important caveat: a strict apples-to-apples out-of-sample RMSE race versus Savant/FanGraphs is not currently published in a directly comparable target/unit format (park-half residual prediction). In this build, we benchmarked against strong internal baselines (zero effect and trailing park mean) and validated invariance against team quality.

## Assumptions and Limits

- Excludes 2020 entirely.
- Park-era correctness depends on `park_era_events.csv`; errors there propagate.
- Weather/drag terms are included when present but are noisy.
- 1H/2H split captures broad seasonal effects, not game-level weather micro-effects.
- New park eras (or one-year samples) remain higher variance despite hierarchical shrinkage.

## Practical Usage for Streaming

Use `Overall Park Factor` as the default environment flag, then adjust for pitcher profile:

- Fly-ball/HR-risk SP: weight `HR Park Factor` more heavily.
- Contact-management/GB SP: weight `BACON Park Factor` more heavily.
- Parks with large HR-BACON divergence should drive matchup-specific overrides.

## Sources

- Baseball Savant Statcast Search CSV endpoint: <https://baseballsavant.mlb.com/statcast_search/csv>
- Baseball Savant OAA leaderboard: <https://baseballsavant.mlb.com/leaderboard/outs_above_average>
- Baseball Savant park factors page: <https://baseballsavant.mlb.com/leaderboard/statcast-park-factors>
- FanGraphs fielding leaders (DRS, UZR): <https://www.fangraphs.com/leaders/major-league?stats=fld>
- MLB team pages for 2026 venue verification: <https://www.mlb.com/team>
- Team/park-era and defense source logs used in build:
  - `data/manual/mlb_home_parks_2026_verified_sources.csv`
  - `data/manual/team_defense_2015_2025_sources.csv`

